Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway

### 2021-04-20 analysis/paper/paper.Rmd
Implications of orthogonal relationship between WABI and core/chip rel freq

```{r get-data, eval = TRUE, results = 'hide'}
assemblages <- read.csv(here::here('analysis/data/raw_data/assemblage_data.csv'))
site_pts <- st_read(here::here('analysis/data/raw_data/sites.gpkg'))
countries <- ne_download(scale = 10, type = 'countries', category = 'cultural',
                         returnclass = "sf")
```

@@

The first part of the analysis involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports. The purpose of this exercise is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier studies that have employed more informally driven methods. This assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviorally meaningful. However, the approach taken is partially informed by the so-called Frison Effect, which highlights the fact that the shape of lithics studied by archaeologists is that which they had when they where discarded - the artefacts can have had long and complex use-lives in which they took on a multitude of different shapes. Several scholars have built on this to argue that morphological variation in retouched lithics is not predominantly the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvination [e.g. @barton1991; @dibble1995] . Consequently, several artefact categories have here been collapased for the CA. This for example pertains to tool types such as scrapers, burins, drills and knives and indeterminate artefacts with retouch. That these categories are internally consistent and categorically exclusive in terms of fulfilled purpose, is at best a dubious proposition, in turn rendering their contribution as discrete analytic units potentially entirely misleading. These have all been combined into the single category "small flint tools". (A full overview of the aggregated variables and their constituent parts is provided in the supplementary material). While aggregating artefact groups in this manner could potentially subsume important variation, it does also reduce the possibility that any drawn conclusions are not simply the result of employing erroneous units of analysis. An underlying assumption is therefore that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes. While ultimately intuitive in nature, it does seem reasonable to assume that for example large non-flint stone tools such as axes, adzes, chisels, clubs and hatches, here categorised as non-flint macro tools, have fulfilled different purposes than the previously mentioned small flint tools.

@@

Figure \@ref(fig:pca-analysis) displays a principle components analysis using variations of the continuous measures for degree of mobility as operationalised by @Bicho2020. While the investigation performed by @Bicho2020 indicates that the relative frequency of chips, cores, and blanks might be more sensitive to mobility patterns, they also found that the relationship between volumetric density of lithics and the frequency of retouched artefacts gave a reasonable estimation of mobility in their data. hese last dimensions are capturing the most variation in the dataset presented here. Thus, if frequency of secondarily worked artefacts is accepted as a proxy for degree of mobility, these findings would consequently be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility than sites from later phases.

@@

The consistent nature of the negative correlation between LVD and relative retouch frequency across this study (r = -0.50), @clark2017 (r = -0.50) and @Bicho2020 (r = -0.48) is striking. While this result is perhaps not surprising given the previous and plentiful indications of the same trends, these results do speak to the cross-cultural and cross-temporal applicability of the measure, and consequently the potential it holds for larger scale comparative studies. The findings indicated by the WABI, as well as by the negative correlation between relative frequency of primarily and secondarily worked lithics, do align more clearly with previous suggestions concerning the overall mobility patterns in Mesolithic Norway. However, the question then becomes: What is the significance of the negatively correlated variables of relative frequency of cores and chips? Given that these are more or less orthogonal to the WABI variables on the first and second component, this would indicate that if one of these variable pairs is taken to reflect the of expedient and curated assemblages, it should follow that the other variable pairs is not correlated with these. As the entire temporal range of the sites is spread out between the extremes of the chip and core variables, one possibility is that these are indicative of site types that are temporally ubiquitous. In attempt to determine if this might be the case, it was decided to introduce locational data from a previous study that in addition to surveyed sites involved the analysis of the sites treated here @roalkvam2020. This found that the most consistently important locational variable for the sites in the region was the exposure of the sites to the elements, and failed to identify any diachronic variation in this pattern. As it has previously been proposed that sites located in areas exposed to wind and waves and with large overview of surrounding areas is a characteristic of hunting stations or similar, one could envisage that variablity in site functionality indicated by the core to chip correlation might be reflected in variation in degree of exposure.

@@

>>>>

```{r get-data, eval = TRUE, results = 'hide'}
# gis <- read.csv(here::here('../../site_data.csv'))
# gis$askeladden_id <- as.integer(str_replace_all(gis$askeladden_id, c("-1" = "", "-0" = "", "-2" = "", "-3" = "", "-4" = "")))

assemblages <- read.csv(here::here('analysis/data/raw_data/assemblage_data.csv'))
site_pts <- st_read(here::here('analysis/data/raw_data/sites.gpkg'))
countries <- ne_download(scale = 10, type = 'countries', category = 'cultural',
                         returnclass = "sf")
```

@@

The first part of the analysis involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports. The purpose of this exercise is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier studies that have employed more informally driven methods. This assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviorally meaningful. However, the approach taken is partially informed by the so-called Frison Effect, which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes, and that the shapes studied by archaeologists are the shape they took upon discard. Several scholars have built on this to argue that morphological variation in retouched lithics is not predominantly the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvination [e.g. @barton1991; @dibble1995]. Consequently, several artefact categories have here been collapased for the CA. This for example pertains to tool types such as scrapers, burins, drills and knives and indeterminate artefacts with retouch. That these categories are internally consistent and categorically exclusive in terms of fulfilled purpose, is at best a dubious proposition, in turn rendering their contribution as discrete analytic units potentially entirely misleading. These have all been combined into the single category "small flint tools". (A full overview of the aggregated variables and their constituent parts is provided in the supplementary material). While aggregating artefact groups in this manner could potentially subsume important variation, it does also reduce the possibility that any drawn conclusions are not simply the result of employing erroneous units of analysis. An underlying assumption is therefore that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes. While ultimately intuitive in nature, it does seem reasonable to assume that for example large non-flint stone tools such as axes, adzes, chisels, clubs and hatches, here categorised as non-flint macro tools, have fulfilled different purposes than the previously mentioned small flint tools.

@@

Figure \@ref(fig:pca-analysis) displays a principle components analysis using variations of the continuous measures for settlement strategies as operationalised by @clark2017 and @Bicho2020. While the investigation performed by @Bicho2020 indicates that the relative frequency of chips, cores, and blanks might be more sensitive to mobility patterns, they also found that the relationship between volumetric density of lithics and the frequency of retouched artefacts gave a reasonable estimation of mobility in their data. hese last dimensions are capturing the most variation in the dataset presented here. Thus, if frequency of secondarily worked artefacts is accepted as a proxy for degree of mobility, these findings would consequently be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility than sites from later phases.

@@

The results of the CA align more or less with results of previous research that has considered the chronological . This would indicate that in general, meaningful patterning is associated with the employed artefact groups.

The consistent nature of the negative correlation between LVD and relative retouch frequency across this study (r = -0.50), @clark2017 (r = -0.50) and @Bicho2020 (r = -0.48) is striking. While this result is perhaps not surprising given the previous and plentiful indications of the same trends, these results do speak to the cross-cultural and cross-temporal applicability of the measure, and consequently the potential it holds for larger scale comparative studies. The findings indicated by the WABI, as well as by the negative correlation between relative frequency of primarily and secondarily worked lithics, do align more clearly with previous suggestions concerning the overall mobility patterns in Mesolithic Norway. However, the question then becomes: What is the significance of the negatively correlated variables of relative frequency of cores and chips? Given that these are more or less orthogonal to the WABI variables on the first and second component, this would indicate that if one of these variable pairs is taken to reflect the scale between expedient and curated technologies, it should follow that the other variable pairs is not correlated with these. As the entire temporal range of the sites is spread out between the extremes of the chip and core variables, one possibility is that these are indicative of site types that are temporally ubiquitous. In attempt to determine if this might be the case, it was decided to introduce site locational data from a previous study that in addition to surveyed sites involved the analysis of the sites treated here @roalkvam2020. Of the variables considered, this study found that the most consistently important environmental covariates for the location of sites in the region was related to the exposure of the sites, and failed to identify any diachronic variation in this pattern. As it has previously been proposed that sites located in areas exposed to wind and waves and with large overview of surrounding areas is a characteristic of hunting stations or similarly temporary , one could envisage that variablity in site functionality might be reflected in degree of exposure. However, introducing these variables does not clarify the picture, as both variables show negligible degrees of correlation with any of the variables pertaining to the assemblage data.

@@

# Conclusion

The results of the CA using the entirety of the lithic assemblages align more or less with results of previous research in south-eastern Norway. This would indicate that in general, meaningful chronological patterning is associated with the employed artefact groups. Precisely what behavioural implication the development in the use of these types have, however, is unclear

The results of considering variables that are have been associated with mobility patterns gave rise to two competing hypotheses. These pertain to whether the degree of curation associated with the assemblages as captured by the changing ratio of primary to secondarily worked lithics over time is the result of changing raw material availability or mobility patterns. The orthogonal relationship between core to chip frequency would in this framework be expected to be capturing the opposite relationship. If degree of curation mainly reflects changes in raw-material quality and availability, this would indicate that mobility patterns are comparable throughout the Mesolithic as the core to chip frequency is not time-dependent. While this would be in line with findings from a recent evaluation of the locational patterns associated with the sites [@roalkvam2020], these findings are the results of an exploratory analysis, and the results presented here are therefore perhaps best seen as potentially laying the framework for later devising a targeted strategic test of these competing hypotheses.
